Natural sources of uncertainty and their implications

### 2017-04-30 book_chapter/paper.tex
biber rollback
reference games section further edits

Linguistic interaction, as captured by a sim-max game, has a rather narrow scope. There is a fixed prior, a fixed utility function and a fixed similarity measure between states. If the sender observes a state $t$ and the receiver guesses state $t'$ the utility of that interaction is proportional to the similarity between $t$ and $t$. Here is a natural example of  a situation captured by a sim-max model: Kiki wants Bubu to draw her dissertation cover in a particular shade of blue $t$; the closer Bubu's color choice matches Kiki's idea, the better. We could say that Kiki wants to \emph{describe} the color for Bubu to pick out. It's a natural thing to want to do with language. But a lot of communication with vague and contiguously variable properties is rather different. In the general spirit of Wittgenstein's method of arranging multiple language games side by side for a more encompassing picture, we should look at some more cases of language use as well. Let us therefore consider a continuum of cases of referential communication.

@@

Interestingly, referential communication can take place under a plethora of different epistemic conditions. A lot of modeling work considers \emph{reference games without displacement} where sender and receiver communicate about a context, i.e. a set of objects, which is immediately accessible in their direct perceptual environment \parencite{BaronchelliGong2010:Modeling-the-Em,Franke2012:Scales-Salience,Franke2012:On-Scales-Salie}.\footnote{add more references later!} Agents may make mistakes about observing exactly how blue or long etc.~a particular object is, but there is no further fundamental uncertainty about the context. This, of course, need not be so. \textcite{lipman_why_2009} and \textcite{Deemter2009:Utility-and-Lan}, for example, consider \emph{reference games with displacement}. Kiki asks Bubu to fetch the red book from the library. No other book than the one Kiki needs will serve. Yet when Kiki describes the book, the perceptual context in which Bubu must make her choice is not co-present to Kiki. Yet other situations are imaginable. In \emph{reference games with partial uncertainty}, for instance, interlocutors may know that they have shared and immediate access only to a subset of the objects in the context, but be uncertain about whether the co-player sees any other objects. This is the set-up for the classic director task in psycholinguistics \parencite[\emph{e.g.}][]{KraussGlucksberg1977:Social-and-nons,Keysar2000,KeyzarLin2003:Limits-on-Theor}.

What would a rational agent do in repeated reference games? Take a universe in which agents play repeated reference games without displacement. A single game has sender and receiver both flawlessly perceive a context of $k$ objects (with $k$ possibly randomly drawn for each encounter); each object has its individual features, e.g., drawn from a rich multidimensional feature space of continuously variable properties \parencite[\emph{e.g.}][]{Franke2012:Scales-Salience,Franke2012:On-Scales-Salie}. The sender chooses a message $m$ for the target object $t$, but she can condition her choice on the current context $c$. Likewise, the receiver chooses a referent $t'$ from the available objects in $c$, again conditional on $c$. But if agents play many such reference games with different contexts $c$ and $c'$, and if strategic rationality means to signal and interpret so as to coordinate on the correct referent, strategic rationality of local choices in $c$ and $c'$ is not enough to guarantee that the agents' beliefs and behavior in $c$ have any relation to their beliefs and behavior in $c'$. In other words, local strategic rationality in a fixed context does not entail a globally consistent behavior across multiple contexts. The situation only gets worse if we want to have agents use the same expressions in a consistent way also across different types of reference games, with possibly different epistemic conditions.

The picture sketched in the last paragraph neglects a rational agent's ability to learn from previous experience and to generalize from prior experience. Which strategies for generalization are best depends on the stochastic properties of the environment in which learning and decision making takes place. The problem for language is that it what works best for on agent depends on what the other agent does. Communication problems are games in the formal sense of delivering payoffs only as a function of what all players do. Consequently, the problem of how best to adapt behavior from past experience in language games also depends on how other agents categorize, generalize and behave. Nothing gets off the ground with hyper-rationality alone. There needs to be learning, evolution and circumstances. As with sim-max games above, biases and other quirks of general cognition are likely only helpful in solving this massive coordination problem.

>>>>

Linguistic interaction, as captured by a sim-max game, has a rather narrow scope. There is a fixed prior, a fixed utility function and a fixed similarity measure between states. 
If the sender observes a state $t$ and the receiver guesses state $t'$ the utility of that interaction is proportional to the similarity between $t$ and $t$. 
Here is a natural example of a situation captured by a sim-max model: Kiki wants Bubu to draw her dissertation cover in a particular shade of blue $t$; the closer Bubu's color choice matches Kiki's idea, the better. Here Kiki wants to \emph{describe} the color for Bubu to pick out or realize from the whole set of possibilities. 
It's a natural thing to want to do with language. 
But a lot of communication with vague and contiguously variable properties is rather different. 
In the general spirit of Wittgenstein's method of arranging multiple language games side by side for a more encompassing picture, we should look at some more cases of language use as well. 
Let us therefore consider a continuum of cases of referential communication.

@@

Interestingly, referential communication can take place under a plethora of different epistemic conditions. 
A lot of modeling work considers \emph{reference games without displacement} where sender and receiver communicate about a context, i.e. a set of objects, which is immediately accessible in their direct perceptual environment \parencite[\emph{e.g.}][]{BaronchelliGong2010:Modeling-the-Em,Franke2012:Scales-Salience,Franke2012:On-Scales-Salie}.
Agents may make mistakes about observing exactly how blue or long etc.~a particular object is, but there is no further fundamental uncertainty about the context. 
This, of course, need not be so. \textcite{lipman_why_2009} and \textcite{Deemter2009:Utility-and-Lan}, for example, consider \emph{reference games with displacement}. 
Kiki asks Bubu to fetch the red book from the library. 
No other book than the one Kiki needs will serve. 
When Kiki describes the book, the perceptual context in which Bubu must make her choice is not co-present to Kiki at the time of speaking. 
Yet other situations are imaginable. 
In \emph{reference games with partial uncertainty}, for instance, interlocutors may know that they have shared and immediate access only to a subset of the objects in the context, but be uncertain about whether the co-player sees any other objects. 
This is the set-up for the classic director task in psycholinguistics \parencite[\emph{e.g.}][]{KraussGlucksberg1977:Social-and-nons,Keysar2000,KeyzarLin2003:Limits-on-Theor}.

What would a rational agent do in repeated reference games? 
Take a universe in which agents play repeated reference games without displacement. 
A single game has sender and receiver both flawlessly perceive a context of $k$ objects (with $k$ possibly randomly drawn for each encounter); each object has its individual features, e.g., drawn from a rich multidimensional feature space of continuously variable properties \parencite[\emph{e.g.}][]{Franke2012:Scales-Salience,Franke2012:On-Scales-Salie}. 
The sender chooses a message $m$ for the target object $t$, but she can condition her choice on the current context $c$. 
Likewise, the receiver chooses a referent $t'$ from the available objects in $c$, again conditional on $c$. 
But if agents play many such reference games with different contexts $c$ and $c'$, and if strategic rationality means to signal and interpret so as to coordinate on the correct referent, strategic rationality of local choices in $c$ and $c'$ is not enough to guarantee that agents' beliefs and behavior in $c$ have any relation to their beliefs and behavior in $c'$. 
An agent could use a certain signal to denote, say, the tallest object in every context with between five and twelve objects or when the number of objects is a prime number, but use the same signal for the darkest object in all other contexts.
This could even be rational if most agents do it this way.
It could evolve as a convention under selection for communicative success as well.
In other words, neither local strategic rationality in a fixed context nor evolutionary adaptation to each single context entails a natural, globally consistent behavior across multiple contexts. 
The situation only gets worse if we want to have agents use the same expressions in a consistent way also across different types of reference games, with possibly different epistemic conditions.

The picture sketched in the last paragraph neglects one crucial thing: learning from observation. 
For rationality, we need to consider a rational agent's ability to learn from previous experience and to generalize from prior experience. 
For evolutionary adaptation, the story so far misses the key fact that languages and practices have to be learnable in appropriate time from the actual observables, lest what is learned and further transmitted will simply be a different language or practice (\mf{insert refs for iterated learning}).
But even if every language learner in history had been perfectly rational, it is unlikely that anyone ever succeeded in reducing all uncertainty about language use in context and say with confidence: this is the threshold that truly separates heaps from non-heaps (\mf{ref / footnote Williamson?}).

 Which strategies for generalization are best depends on the stochastic properties of the environment in which learning and decision making takes place. The problem for language is that it what works best for on agent depends on what the other agent does. Communication problems are games in the formal sense of delivering payoffs only as a function of what all players do. Consequently, the problem of how best to adapt behavior from past experience in language games also depends on how other agents categorize, generalize and behave. Nothing gets off the ground with hyper-rationality alone. There needs to be learning, evolution and circumstances. As with sim-max games above, biases and other quirks of general cognition are likely only helpful in solving this massive coordination problem. Maybe our world is a place where particular learning biases are needed for any regular and shared communicative practice to get off the ground.
